Component response rate variation drives stability in large complex systems

### 2018-07-30 notebook/SI.Rmd
Final touch ups

**This supplemental information supports the manuscript "Component response rate variation drives stability in large complex systems" with additional analyses to support its conclusions. All text, code, and data underlying this manuscript are publicly available on [GitHub](https://github.com/bradduthie/RandomMatrixStability) as part of the RandomMatrixStability package.**

@@

While the foundational work of May[@May1972] applies broadly to complex networks, much attention has been given specifically to ecological networks of interacting species. In these networks, the matrix $\mathbf{A}$ is interpreted as a community matrix and each row and column is interpreted as a single species. The effect that the density of any species $i$ has on the population dynamics of species $j$ is found in $A_{ij}$, meaning that $\mathbf{A}$ holds the effects of pair-wise interactions between $S$ species[@Allesina2012; @Allesina2015]. While May's original work[@May1972] considered only randomly assembled communities, recent work has specifically looked at more restricted ecological communities including competitive networks (all off-diagonal elements of $\mathbf{A}$ are negative), mutualist networks (all off-diagonal elements of $\mathbf{A}$ are positive), and predator-prey networks (for any pair of $i$ and $j$, the effect of $i$ on $j$ is negative and $j$ on $i$ is positive, or vice versa)[@Allesina2012; @Allesina2015]. In general, competitor and mutualist networks tend to be unstable, while predator-prey networks tend to be highly stabilising.

@@

In the main text, for simplicity, I assumed connectance values of $C = 1$, meaning that all off-diagonal elements of a matrix $\mathbf{M}$ were potentially nonzero and sampled from a normal distribution $\mathcal{N}(0, \sigma^{2})$ where $\sigma = 0.4$. Here I present four tables showing the number of stable communities given $C = \{0.3, 0. 5, 0.7, 0.9 \}$. In all cases, uniform variation in component response time ($\gamma \sim \mathcal{U}(0, 2)$) led to a higher number of stable communities than when $\gamma$ did not vary ($\gamma = 1$). In contrast to the main text, 100000 rather than 1 million $\mathbf{M}$ were simulated. As with the results on [stability with increasing $S$](#IncrS) shown above, in the tables below `A0` refers to matrices when $\gamma = 1$, and `A1` refers to matrices after $Var(\gamma)$ is added. The column `A0_unstable` shows the number of `A0` matrices that are unstable, and the column `A0_stable` shows the number of `A0` matrices that are stable (these two columns sum to 100000). Similarly, the column `A1_unstable` shows the number of `A1` matrices that are unstable and `A1_stable` shows the number that are stable. The columns `A1_stabilised` and `A1_destabilised` show how many `A0` matrices were stabilised or destabilised, respectively, by $Var(\gamma)$.

>>>>

**This supplemental information supports the manuscript "Component response rate variation drives stability in large complex systems" with additional analyses to support its conclusions. All text, code, and data underlying this manuscript are publicly available on [GitHub](https://github.com/bradduthie/RandomMatrixStability) as part of the RandomMatrixStability R package.**

@@

While the foundational work of May[@May1972] applies broadly to complex networks, much attention has been given specifically to ecological networks of interacting species. In these networks, the matrix $\mathbf{A}$ is interpreted as a community matrix and each row and column is interpreted as a single species. The per capita effect that the density of any species $i$ has on the population dynamics of species $j$ is found in $A_{ij}$, meaning that $\mathbf{A}$ holds the effects of pair-wise interactions between $S$ species[@Allesina2012; @Allesina2015]. While May's original work[@May1972] considered only randomly assembled communities, recent work has specifically looked at more restricted ecological communities including competitive networks (all off-diagonal elements of $\mathbf{A}$ are negative), mutualist networks (all off-diagonal elements of $\mathbf{A}$ are positive), and predator-prey networks (for any pair of $i$ and $j$, the effect of $i$ on $j$ is negative and $j$ on $i$ is positive, or vice versa)[@Allesina2012; @Allesina2015]. In general, competitor and mutualist networks tend to be unstable, while predator-prey networks tend to be highly stabilising[@Allesina2012].

@@

In the main text, for simplicity, I assumed connectance values of $C = 1$, meaning that all off-diagonal elements of a matrix $\mathbf{M}$ were potentially nonzero and sampled from a normal distribution $\mathcal{N}(0, \sigma^{2})$ where $\sigma = 0.4$. Here I present four tables showing the number of stable communities given $C = \{0.3, 0. 5, 0.7, 0.9 \}$. In all cases, uniform variation in component response rate ($\gamma \sim \mathcal{U}(0, 2)$) led to a higher number of stable communities than when $\gamma$ did not vary ($\gamma = 1$). In contrast to the main text, 100000 rather than 1 million $\mathbf{M}$ were simulated. As with the results on [stability with increasing $S$](#IncrS) shown above, in the tables below `A0` refers to matrices when $\gamma = 1$, and `A1` refers to matrices after $Var(\gamma)$ is added. The column `A0_unstable` shows the number of `A0` matrices that are unstable, and the column `A0_stable` shows the number of `A0` matrices that are stable (these two columns sum to 100000). Similarly, the column `A1_unstable` shows the number of `A1` matrices that are unstable and `A1_stable` shows the number that are stable. The columns `A1_stabilised` and `A1_destabilised` show how many `A0` matrices were stabilised or destabilised, respectively, by $Var(\gamma)$.

### 2018-07-30 notebook/SI.tex
Final touch ups

\textbf{This supplemental information supports the manuscript
``Component response rate variation drives stability in large complex
systems'' with additional analyses to support its conclusions. All text,
code, and data underlying this manuscript are publicly available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub} as
part of the RandomMatrixStability package.}

@@

While the foundational work of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} applies broadly
to complex networks, much attention has been given specifically to
ecological networks of interacting species. In these networks, the
matrix \(\mathbf{A}\) is interpreted as a community matrix and each row
and column is interpreted as a single species. The effect that the
density of any species \(i\) has on the population dynamics of species
\(j\) is found in \(A_{ij}\), meaning that \(\mathbf{A}\) holds the
effects of pair-wise interactions between \(S\)
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{3}}.
While May's original
work\textsuperscript{\protect\hyperlink{ref-May1972}{1}} considered only
randomly assembled communities, recent work has specifically looked at
more restricted ecological communities including competitive networks
(all off-diagonal elements of \(\mathbf{A}\) are negative), mutualist
networks (all off-diagonal elements of \(\mathbf{A}\) are positive), and
predator-prey networks (for any pair of \(i\) and \(j\), the effect of
\(i\) on \(j\) is negative and \(j\) on \(i\) is positive, or vice
versa)\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{3}}.
In general, competitor and mutualist networks tend to be unstable, while
predator-prey networks tend to be highly stabilising.

@@

In the main text, for simplicity, I assumed connectance values of
\(C = 1\), meaning that all off-diagonal elements of a matrix
\(\mathbf{M}\) were potentially nonzero and sampled from a normal
distribution \(\mathcal{N}(0, \sigma^{2})\) where \(\sigma = 0.4\). Here
I present four tables showing the number of stable communities given
\(C = \{0.3, 0. 5, 0.7, 0.9 \}\). In all cases, uniform variation in
component response time (\(\gamma \sim \mathcal{U}(0, 2)\)) led to a
higher number of stable communities than when \(\gamma\) did not vary
(\(\gamma = 1\)). In contrast to the main text, 100000 rather than 1
million \(\mathbf{M}\) were simulated. As with the results on
\protect\hyperlink{IncrS}{stability with increasing \(S\)} shown above,
in the tables below \texttt{A0} refers to matrices when \(\gamma = 1\),
and \texttt{A1} refers to matrices after \(Var(\gamma)\) is added. The
column \texttt{A0\_unstable} shows the number of \texttt{A0} matrices
that are unstable, and the column \texttt{A0\_stable} shows the number
of \texttt{A0} matrices that are stable (these two columns sum to
100000). Similarly, the column \texttt{A1\_unstable} shows the number of
\texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows the
number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

>>>>

\textbf{This supplemental information supports the manuscript
``Component response rate variation drives stability in large complex
systems'' with additional analyses to support its conclusions. All text,
code, and data underlying this manuscript are publicly available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub} as
part of the RandomMatrixStability R package.}

@@

While the foundational work of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} applies broadly
to complex networks, much attention has been given specifically to
ecological networks of interacting species. In these networks, the
matrix \(\mathbf{A}\) is interpreted as a community matrix and each row
and column is interpreted as a single species. The per capita effect
that the density of any species \(i\) has on the population dynamics of
species \(j\) is found in \(A_{ij}\), meaning that \(\mathbf{A}\) holds
the effects of pair-wise interactions between \(S\)
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{3}}.
While May's original
work\textsuperscript{\protect\hyperlink{ref-May1972}{1}} considered only
randomly assembled communities, recent work has specifically looked at
more restricted ecological communities including competitive networks
(all off-diagonal elements of \(\mathbf{A}\) are negative), mutualist
networks (all off-diagonal elements of \(\mathbf{A}\) are positive), and
predator-prey networks (for any pair of \(i\) and \(j\), the effect of
\(i\) on \(j\) is negative and \(j\) on \(i\) is positive, or vice
versa)\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{3}}.
In general, competitor and mutualist networks tend to be unstable, while
predator-prey networks tend to be highly
stabilising\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.

@@

In the main text, for simplicity, I assumed connectance values of
\(C = 1\), meaning that all off-diagonal elements of a matrix
\(\mathbf{M}\) were potentially nonzero and sampled from a normal
distribution \(\mathcal{N}(0, \sigma^{2})\) where \(\sigma = 0.4\). Here
I present four tables showing the number of stable communities given
\(C = \{0.3, 0. 5, 0.7, 0.9 \}\). In all cases, uniform variation in
component response rate (\(\gamma \sim \mathcal{U}(0, 2)\)) led to a
higher number of stable communities than when \(\gamma\) did not vary
(\(\gamma = 1\)). In contrast to the main text, 100000 rather than 1
million \(\mathbf{M}\) were simulated. As with the results on
\protect\hyperlink{IncrS}{stability with increasing \(S\)} shown above,
in the tables below \texttt{A0} refers to matrices when \(\gamma = 1\),
and \texttt{A1} refers to matrices after \(Var(\gamma)\) is added. The
column \texttt{A0\_unstable} shows the number of \texttt{A0} matrices
that are unstable, and the column \texttt{A0\_stable} shows the number
of \texttt{A0} matrices that are stable (these two columns sum to
100000). Similarly, the column \texttt{A1\_unstable} shows the number of
\texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows the
number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

### 2018-07-30 notebook/ms.Rmd
Final touch ups

**The stability of a complex system generally decreases with increasing system size and interconnectivity, a counter-intuitive result of widespread importance across the physical, life, and social sciences. Despite recent interest in the relationship between system properties and stability, the effect of variation in the response rate of individual system components remains unconsidered. Here I vary the component response rates ($\boldsymbol{\gamma}$) of randomly generated complex systems. I show that when component response rates vary, the potential for system stability is markedly increased. Variation in $\boldsymbol{\gamma}$ becomes increasingly important as system size increases, such that the largest stable complex systems would be unstable if not for $\boldsymbol{Var(\gamma)}$. My results reveal a previously unconsidered driver of system stability that is likely to be pervasive across all complex systems.**

@@

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that individual components respond to perturbations of the system at the same rate ($\gamma$), but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of individuals, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, complex systems in which $\gamma$ varies are actually more likely to be stable, especially when $S$ is high.

@@

**Component response rates of random complex systems**. Rows in $\mathbf{M}$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a real scalar value $\gamma_{i}$ [@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $\mathbf{M}$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $\mathbf{M}$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn randomly from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $\gamma_{i}$ to generate $\mathbf{M}$. When $\gamma_{i} = 1$, eigenvalues of $\mathbf{M}$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (red circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $\mathbf{M}$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

**Simulation of random $\mathbf{M}$ across $\mathbf{S}$**. To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $\mathbf{M}$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from 2-50. One million $\mathbf{M}$ were simulated for each $S$, and the stability of $\mathbf{M}$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$. For all $S > 10$, I found that the number of stable random systems was higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3; see Supplementary Information for full table of results), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$.

@@

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibile equilibria assume that the values of all system components are positive at equilibrium[@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic[@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities[@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). Feasibility was unaffected by $Var(\gamma)$, meaning that for pure interacting species networks, variation in component response time (i.e., species generation time) does not affect stability at biologically realistic species densities.

@@

Here I have shown that the stability of large systems might be often contigent upon variation in the response rates of their individual components, meaning that factors such as generation time and rate of trait evolution (in biological networks), transaction speed (in economic networks), or communication speed (in social networks) need to be considered when investigating the stability of complex systems. Variation in component response rate becomes more likely to be critical for stability as system size increases, and can ultimately increase the overall probability that system stability is observed above that predicted by May's[@May1972] classically derived $\sigma \sqrt{SC}$ criterion. The logic outlined here is general, and potentially applies to any complex system in which individual system components can vary in their reaction rates to system perturbation.

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. The mathematics underlying this shift in eigenvalue distribution has been investigated[@Ahmadian2015] and recently applied to questions concerning species density and feasibility[@Gibbs2017; @Stone2017], but has not been interpreted as rates of response of individual componetents to system perturbation.

The potential importance of component response rate variation was most evident from the results of simulations in which the genetic algorithm was used in attempt to maximise the probability of system stability. The probability that some combination of component respose rates could be found to stabilise the system was shown to be up to four orders of magnitude higher than the background probabilities of stability in the absence of any component response rate variation. Instead of manipulating the $S \times S$ interactions between system components, it might therefore be possible to manipulate only the $S$ response rates of individual system components to achieve stability. Hence, managing the response rates of system components in a targeted way could potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

Interestingly, while complex systems were more likely to be stable given variation in component respose rate, they were not more likely to be feasible, meaning that stability was not increased when component values were also restricted to being positive at equilibrium. Feasibility is important to consider, particularly for the study of ecological networks of species[@Stone2017; @Grilli2017; @Dougoud2018] because population densities cannot realistically be negative. My results therefore suggest that variation in the rate of population responses to perturbation (e.g., due to differences in generation time among species) is unlikely to be critical to the stability of purely multi-species interaction networks (see also Supplementary Information). Nevertheless, ecological interactions do not exist in isolation in empirical systems[@Patel2018], but instead interact with evolutionary, abiotic, or social-economic systems. The relevance of component response rate for complex system stability should therefore not be ignored in the broader context of ecological communities.

@@

In the above, $\mathbf{x^{*}}$ is the vector of species abundances at equilibrium. Feasibility is satisfied if all elements in $\mathbf{x^{*}}$ are positive. The matrix $\mathbf{I}$ is the identity matrix, and the value $\theta$ is strength of intraspecific competition (diagonal elements). Diagonal values are set to $-1$, so $\theta = -1$. The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$) for $\mathbf{J}$. Implicitly, here $\delta = 0$ underlying strong interactions. Hence, $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta \mathbf{I}$) is added to $\mathbf{J}$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's[@May1972] stability criterion[@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1$, and hence for my purposes $(CS)^{-\delta} = 1$). Given $\mathbf{A} = \theta\mathbf{I + J}$, the above criteria is therefore reduced to the below,

>>>>

**The stability of a complex system generally decreases with increasing system size and interconnectivity, a counterintuitive result of widespread importance across the physical, life, and social sciences. Despite recent interest in the relationship between system properties and stability, the effect of variation in the response rate of individual system components remains unconsidered. Here I vary the component response rates ($\boldsymbol{\gamma}$) of randomly generated complex systems. I show that when component response rates vary, the potential for system stability is markedly increased. Variation in $\boldsymbol{\gamma}$ becomes increasingly important as system size increases, such that the largest stable complex systems would be unstable if not for $\boldsymbol{Var(\gamma)}$. My results reveal a previously unconsidered driver of system stability that is likely to be pervasive across all complex systems.**

@@

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that individual components respond to perturbations of the system at the same rate ($\gamma$), but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, complex systems in which $\gamma$ varies are actually more likely to be stable, especially when $S$ is high.

@@

**Component response rates of random complex systems**. Rows in $\mathbf{M}$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a real scalar value $\gamma_{i}$ [@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $\mathbf{M}$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $\mathbf{M}$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn randomly from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $\gamma_{i}$ to generate $\mathbf{M}$. When $\gamma_{i} = 1$, eigenvalues of $\mathbf{M}$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (red circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$), while the latter cluster has a smaller radius, but real components have shifted to the right; all $\Re({\lambda}) < 0$. Overall, for 1 million randomly assembled $\mathbf{M}$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

**Simulation of random $\mathbf{M}$ across $\mathbf{S}$**. To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $\mathbf{M}$ matrices at $\sigma = 0.4$ and $C = 1$ across $S = \{2, 3, ..., 49, 50\}$. One million $\mathbf{M}$ were simulated for each $S$, and the stability of $\mathbf{M}$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$. For all $S > 10$, I found that the number of stable random systems was higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3; see Supplementary Information for full table of results), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$.

@@

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that the system is stable and values of all components are positive at equilibrium[@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic[@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities[@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). Feasibility was unaffected by $Var(\gamma)$, meaning that for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.

@@

Here I have shown that the stability of large systems might often be contigent upon variation in the response rates of their individual components, meaning that factors such as rate of trait evolution (in biological networks), transaction speed (in economic networks), or communication speed (in social networks) need to be considered when investigating the stability of complex systems. Variation in component response rate becomes more likely to be critical for stability as system size increases, and can ultimately increase the overall probability that system stability is observed above that predicted by May's[@May1972] classically derived $\sigma \sqrt{SC}$ criterion. The logic outlined here is general, and potentially applies to any complex system in which individual system components can vary in their reaction rates to system perturbation.

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. The mathematics underlying this shift in eigenvalue distribution has been investigated[@Ahmadian2015] and recently applied to questions concerning species density and feasibility[@Gibbs2017; @Stone2017], but has not been interpreted as rates of response of individual system components to perturbation.

The potential importance of component response rate variation was most evident from the results of simulations in which the genetic algorithm was used in attempt to maximise the probability of system stability. The probability that some combination of component response rates could be found to stabilise the system was shown to be up to four orders of magnitude higher than the background probabilities of stability in the absence of any component response rate variation. Instead of manipulating the $S \times S$ interactions between system components, it might therefore be possible to manipulate only the $S$ response rates of individual system components to achieve stability. Hence, managing the response rates of system components in a targeted way could potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

Interestingly, while complex systems were more likely to be stable given variation in component response rate, they were not more likely to be feasible, meaning that stability was not increased when component values were also restricted to being positive at equilibrium. Feasibility is important to consider, particularly for the study of ecological networks of species[@Stone2017; @Grilli2017; @Dougoud2018] because population densities cannot realistically be negative. My results therefore suggest that variation in the rate of population responses to perturbation (e.g., due to differences in generation time among species) is unlikely to be critical to the stability of purely multi-species interaction networks (see also Supplementary Information). Nevertheless, ecological interactions do not exist in isolation in empirical systems[@Patel2018], but instead interact with evolutionary, abiotic, or social-economic systems. The relevance of component response rate for complex system stability should therefore not be ignored in the broader context of ecological communities.

@@

In the above, $\mathbf{x^{*}}$ is the vector of species densities at equilibrium. Feasibility is satisfied if all elements in $\mathbf{x^{*}}$ are positive. The matrix $\mathbf{I}$ is the identity matrix, and the value $\theta$ is the strength of intraspecific competition (diagonal elements). Diagonal values are set to $-1$, so $\theta = -1$. The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$) for $\mathbf{J}$. Implicitly, here $\delta = 0$ underlying strong interactions. Hence, $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta \mathbf{I}$) is added to $\mathbf{J}$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's[@May1972] stability criterion[@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1$, and hence for my purposes $(CS)^{-\delta} = 1$). Given $\mathbf{A} = \theta\mathbf{I + J}$, the above criteria is therefore reduced to the below,
